Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-06 talk/pepm2011/paper.tex
change title, add michael as an author, refactor some small things

\title{Escape Analysis and Specialization in a Tracing JIT}

\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciej Fijałkowski \and Samuele Pedroni \and Armin Rigo}
           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
           {cfbolz@gmx.de, antocuni@gmail.com, fijal@merlinux.eu,
           samuele.pedroni@gmail.com, arigo@tunes.org}

@@

\section{Escape Analysis in a Tracing JIT}
\label{sec:virtuals}

@@

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object\arigo{XXX what I have in mind when I talk
of ``virtual object'' is the run-time behavior -- i.e. a real object that
would exist at run-time, except that it has be virtual-ized.  Here you seem
to mean rather ``virtual object description'' or something.}
is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the optimization to improve the trace. The shapes describe
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.
Equivalently, a \texttt{guard\_class} on a virtual object can be removed as
well, because the virtual object has a fixed and known class.

In the example from last section, the following operations would produce two
virtual objects, and be completely removed from the optimized trace:

@@

The virtual object stored in $p_{5}$ would know that it is an \texttt{BoxedInteger}, and that
the \texttt{intval} field contains $i_{4}$, the one stored in $p_{6}$ would know that
its \texttt{intval} field contains the constant -100.

@@

So far we have only described what happens when virtual objects are used in
operations that read and write their fields and in guards. When the virtual
object is used in
any other operation, it cannot stay virtual. For example, when a virtual object
is stored in a globally accessible place, the object needs to actually be
allocated, as it might live longer than one iteration of the loop.

@@

\subsection{Algorithm}
\label{sub:Algorithm}

@@

% subsection Algorithm (end)

>>>>

\title{Allocation Removal by Partial Evaluation in a Tracing JIT}

\authorinfo{Carl Friedrich Bolz \and Antonio Cuni \and Maciej Fijałkowski \and Michael Leuschel \and Samuele Pedroni \and Armin Rigo}
           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
           {cfbolz@gmx.de, anto.cuni@gmail.com, fijal@merlinux.eu,
           samuele.pedroni@gmail.com, arigo@tunes.org}

@@

\section{Escape Analysis in a Tracing JIT} % XXX imprecise title
\label{sec:virtuals}

@@

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a shape description is constructed and associated with the variable
that would have stored the result of \texttt{new}. This result, a newly
allocated object, is called \emph{virtual}, because after the optimization is
finished, it does not need to be allocated at all. The shape description summarizes
the shape of this original object
and is used by the optimization to improve the trace. The shape describes
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into such
an object, that shape description is updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from such an object, the result is read
from the shape description, and the operation is also removed.
Equivalently, a \texttt{guard\_class} on a variable that has a shape description can be removed as
well, because the shape description stores the class.

In the example from last section, the following operations would produce two
shape descriptions, and be completely removed from the optimized trace:

@@

The shape description associated with $p_{5}$ would know that it is an
\texttt{BoxedInteger}, and that the \texttt{intval} field contains $i_{4}$, the
associated with $p_{6}$ would know that its \texttt{intval} field contains the
constant -100.

@@

So far we have only described what happens when virtual objects are used in
operations that read and write their fields and in guards. When the virtual
object is used in
any other operation, it cannot stay virtual. For example, when a virtual object
is stored in a globally accessible place, the object needs to actually be
allocated, as it might live longer than one iteration of the loop. The actual
allocation is inserted just before the operation that requires the object to be
non-virtual.

@@

\section{Formal Description of the Algorithm}
\label{sec:Algorithm}

@@

% section Formal Description of the Algorithm (end)
